Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability

### 2023-02-02 paper.tex
Edits to results section.

\subsection{Effects of changes to system granularity}
\begin{figure*}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{GranularityPerturbation.png}
\vspace{-0.1in}
\caption{Effects of Scale Changes. (A-F) Joint (A,C,D,) and marginal (B,D,F) probability distributions for the FAC control method at $\alpha=0.2$ (A,B), $\alpha=1$ (C,D) and $\alpha=2.0$ (E,F).  Target point shown as blue circle. (G-L) Same as (A-F), but for PAC controller. (M)  FAC (cyan) and PAC (magenta) controller performance as function of system volume. Horizontal dashed lines correspond to SSA analyses at $\alpha=100$. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021} other than their adaptation to the new system volume size (see Methods).}
\label{Volume}
\end{center}
\vspace{-0.2in}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{TimeDelayPerturbation.png}
\vspace{-0.1in}
\caption{Effects of time delays. (Ai-Fi) Joint (Ai,Ci,Di) and marginal (Bi,Di,Fi) probability distributions for the FAC control method at three different time delays of $\tau/\gamma$=0.01 (top), 0.1 (middle) and 1 (bottom) relative to the target point (blue circles). (Aii-Fii) Same as (Ai-Fi), but for PAC controller. (G) FAC PAC performance as function of time delay show that PAC outperforms the FAC when time delay is greater than $0.07 /\gamma$. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}.}
\label{Time}
\end{center}
\vspace{-0.2in}
\end{figure*}

@@

Time delay was hypothesized to adversely affect control performance of the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers. To test the effect of time delay on control performance, groups of time delayed SSA were simulated while changing controller and the level of time delay between each group. Each SSA was sampled for 1000 times over 10000 minutes after reaching a steady state by simulating the system for 10000 minutes.

Figure\ \ref{Time} shows the joint and marginal of each controller and time-delay pair and the resulting score, with panels Ai - Fi showing results for the FAC controller and panels Aii-Fii showing results for the PAC controller, and panel G showing the score of both controllers versus the time delay. From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale of the process. At low time delays (below $\tau = 0.07/\gamma$), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma$) but at moderate time delays (above $\tau = 0.07/\gamma$) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma$)).

This data taken together show that time delays much larger that $0.07 / \gamma$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.  For extreme levels of time delay, both systems lose their asymmetry, and their scores become much worse ($J=$1041 and 931).

\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.png}
\caption{Effects of parametric variations to Cell 1 (top row), Cell 2 (middle row) or to both cells simultaneously (bottom row). Each column show the effect of changing a single model parameter (e.g., $\kappa$ in the leftmost row). Trends for the FAC (solid lines) and PAC (dashed lines), but the optimal controller can change under parameter variations. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}.}
\label{Parameter}
\end{center}
\end{figure*}

It is rare that any parameters are exactly known for biological processes, and real systems often exhibit slightly different parameters due to extrinsic variations. This parameter variation between cells is factor that could influence control performance. To test this, parameter sensitivity analysis was performed on the cell one while holding the parameters of cell two constant figure \ref{Parameter}(Row I). For each perturbation analysis, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controller was used over a range of $s$ between 0.5 and 2. The data for such perturbations are shown in figure \ref{Parameter}. The analysis was then repeated, this time holding the parameters of Cell 1 constant and perturbing only the parameters of Cell 2  (shown in figure\ \ref{Parameter}(Row II)). Finally, both cell one and cell two had their parameters jointly perturbed by the same amount (figure \ref{Parameter}(Row III)).

Parameter changes were found to have a broad range of effects, including increasing performance, decreasing performance, or having little effect. Figure \ref{Parameter}(Bi) shows that increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance. In some cases, these effects were not monotonic; for example increasing $\kappa$ in Cell 1 is highly advantageous up to a limit after which the control performance degrades rapidly. In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations. When parameters of both Cell 1 and Cell 2 are jointly changed (Fig.\ \ref{Parameter} bottom row) some unique trends were observed. It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).

When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which caused a decrease in the control performance of the FAC also tended to decrease the performance of the PAC. Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.  One explanation for why changes to $\eta$ had no effect on control performance is that the origonal parameter was so small that it may require a larger scaling on a different magnitude before changes are observed. For situations where the $\mathbf{u}^{\rm PAC}$ outperformed the $\mathbf{u}^{\rm FAC}$ at large $s$, this may be an indication that re-optimizing the controller for this parameter combination could yield better control results, since the controller is not re-optimized at each parameter combination.

>>>>

\subsection{Effects of changes to system granularity}
\begin{figure*}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{GranularityPerturbation.png}
\vspace{-0.1in}
\caption{Effects of Scale Changes. (A-F) Joint (A,C,D,) and marginal (B,D,F) probability distributions for the FAC control method at $\alpha=0.2$ (A,B), $\alpha=1$ (C,D) and $\alpha=2.0$ (E,F).  Target point shown as blue circle. (G-L) Same as (A-F), but for PAC controller. (M)  FAC (cyan) and PAC (magenta) controller performance as function of system volume. Horizontal dashed lines correspond to SSA analyses at $\alpha=100$. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021} other than their adaptation to the new system volume size (see Methods).}
\label{Volume}
\end{center}
\vspace{-0.2in}
\end{figure*}
% XXX - This figure needs larger fonts for all the labels and axes. 
% Update panel letters as written in caption
% Add tic marks to left side of the A and C heat maps

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.png}
\caption{Effects of parametric variations on control performance.  ({\bf A}) Sensitivity of control performance score, $J$, to modifications in parameters applied to Cell 1 only, one at a time for parameters $\kappa$, XXX, XXX, XXX, and XXX (in order left to right).  ({\bf B}) Same as {\bf A}, but where modifications are made to the corresponding parameter in Cell 2. ({\bf C}) Same as ({\bf A}), but where the modification is made simultaneously to both cells. In all panels, trends lines are shown for the FAC (solid) and PAC (dashed) controllers. Controllers are unchanged from \cite{May2021}.}
\label{Parameter}
\end{center}
\end{figure*}
% XXX - This figure needs much larger fonts for all the labels and axes.
% There are errors in the parameter names.  What is 'alpha'?
% remove the A,B,C,D,E.
% replace I, II, III with A, B, C
% what is parameter $\mu$? Is that an error?  Please use $\mu$ for reaction index as in the methods section.

Exact parameters are never known for any real biological process, and even genetically identical single-cell systems are expected to exhibit cell-to-cell heterogeneity in parameters due to extrinsic variations outside of the specific pathways of interest. These parameter uncertainties or variations between cells could influence system performance, such that a control law that is optimized for one fixed, but incorrect, set of parameters may or may not succeed if used to break symmetry and track a desired target state when applied to a system with different parameters. To explore the sensitivity of control performance to unknown errors in the model parameters, sensitivity analysis was performed for each model parameter and for the situation where the corresponding parameter was modified in either one cell at a time (i.e., to simulate extrinsic noise) or in both cells simultaneously (i.e., to simulate parameter errors).  For each perturbation analysis, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controller were analyzed over a range of parameters from one-half to two-fold of the original parameters.
% XXX - How hard would it be to extend this to 10 fold in either direction?
 Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is change while the parameters of Cell 2 is held fixed, and Fig.\ \ref{Parameter}B shows the opposite when changes Cell 1 has fixed parameters and Cell 2 is modified.  Finally, \ref{Parameter}B  shows the performance change when the corresponding parameter is changed in both Cell 1 and Cell 2 at the same time.

Modification to parameters was found to produce a broad range of effects, where some changes increase performance, some decrease performance, and others have little or no effect. For example, increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Fig.\ \ref{Parameter}A and \ref{Parameter}B). In some cases, these effects were not monotonic; for example, increasing $\kappa$ in Cell 1 (Fig.\ \ref{Parameter}A, left column) would be highly advantageous up to a limit after which the control performance degrades rapidly. In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations.Similarly, Fig.\ \ref{Parameter}C shows that when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance.
%It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).

When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which cause a decrease in the control performance of the FAC also tend to decrease the performance of the PAC. Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.  One explanation for why changes to $\eta$ had no effect on control performance is that the original parameter was so small that it may require a larger scaling on a different magnitude before changes are observed. 
% XXX - what is \eta?  From the table this should be the cooperativity with is very large.  Is this a mistake?
In a few circumstances, changes in the parameters $\kappa$ or $\gamma$ led to the surprising situation where the PAC controller, which only observes one cell at a time, was more robust to parameter errors than was the FAC controller, which can observe both cells, even to the extent that the PAC controller could sometimes outperform the FAC.  We stress that this effect is due to the mismatch between the assumed and the actual parameters, and re-optimizing the two controllers for the new parameter combination will always yield better control results for the FAC compared to the PAC.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{TimeDelayPerturbation.png}
\vspace{-0.1in}
\caption{Effects of time delays on control performance. (A-F) Joint (A,C,E) and marginal (B,D,F) probability distributions for the FAC control method at three different time delays of $\tau/\gamma$=0.01 (A,B), 0.1 (C,D) and 1.0 (E,F) relative to the target point (blue circles). (G-L) Same as (A-F), but for PAC controller. (M) FAC and PAC performance as function of time delay show that PAC outperforms the FAC when time delay is greater than $0.07 /\gamma$. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}; specifically, the protein degradation rate is $\gamma = 0.0203$ as identified from \cite{Baumchager}.}
\label{Time}
\end{center}
\vspace{-0.2in}
\end{figure*}
% XXX - This figure needs larger fonts for all the labels and axes. 
% Update panel letters as written in caption
% Add tic marks to left side of the A and C heat maps
% Fix the ref to baumschlager paper.

Feedback control can only be effective if one can quickly make measurements, compute adjustment to the control signal, and implement the needed change to the system. As the time required for any of these steps increases, control performance will be degraded, perhaps even leading to large fluctuations or instability. To explore how time delays would affect the noise-enhance controller $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$, we generated large sets of time-delayed stochastic simulations (see Methods) for different lengths of the time delay. Each SSA was sub-sampled for 1000 times over 10000 minutes of simulation time after a burnin period of 10000 minutes.

Figure \ref{Time} shows the joint and marginal distribution of each controller and time-delay pair and the resulting score, with panels A-F showing results for the FAC controller and panels G-L showing results for the PAC controller. Figure \ref{Time}M summarizes these results by plotting the score of both controllers versus the time delay. From the figures, it is clear that performance is rapidly degraded as the delay approaches and then exceeds the characteristic time scale of the process. At ver small time delays (below $\tau = 0.07/\gamma = XXX$ min), the FAC outperforms the PAC ($J=$87 versus 146 at $\tau = 0.01/\gamma = XXX$ min) but at moderate time delays (above $\tau = 0.07/\gamma = XXX$ min) the PAC outperforms the FAC ($J=$170 versus 219 at $\tau = 0.1/\gamma = XXX$ min). 
% XXX - you can use the value of gamma=0.203 in Table 1 to write these in units of minutes.

This data taken together show that time much larger that $0.07 / \gamma = XXX min$ is detrimental to the FAC controller; delays beyond $0.2 / \gamma = XXX min$ are detrimental for the PAC controller; and the best choice in controller is dependent the level of time delay in the system.  For extreme levels of time delay, both systems lose their asymmetry, and their scores become much worse ($J=$1041 and 931).
